---
title: "Cleaning - Week 3"
author: "Syed Arif Hussain"
date: '2022-05-09'
output: html_document
---

## Subsetting

- Subset a column: X[,1]
- Subset by column: X[, "varname"]

- Subset by row and column: X[1:2, "varname]

- Subset using logic:
    - X[(X$var1 <= 3 & X$var2 > 11), ] --> gives only data on the rows that fulfill this statement
    - X[(X$var1 <= 3 | X$var2 >15), ] --> or

Dealing with NA values
- Subsetting on NA does not produce the row --> use which command
    - Returns indices 
    - e.g., X[which(X$var2 > 8),] --> returns indices in which var2 > 8
        - Does not return the NA

```{r }

```

## Sorting Variables
- Sort using sort() command
- sort(x$var1) --> in increasing order; 
- decreasing order --> sort(, decreasing=TRUE)
- dealing with NA --> sort(,na.last = TRUE)

Order df by a variable
- x[order(X$var1), ] --> will order the values by variable 1 in increasing order
    - this is reodering the rows which is why we are passing it through row index 

- Can order by multiple variables:
  - x[order(x$var1, x$var2), ]
  - Sorts first variable first and then second within first

```{r}
```

#Sorting in plyr package
- use the arrange command --> arrange(dataframe, var1) --> sorts df on variable 1
    - same as order
- arrange(dataframe, desc(var1)) --> decreasing order

#Adding rows and columns to df
- x$var4 <- rnowm(5) --> var4 is a variable not in df; we assigned rnorm(5) to it which adds it

- Can also add with cbind:
    - Y <- cbind(dataframe, norm(5)) --> adds a column bind to dataframe on right side
    - Y <- cbind(norm(5),dataframe) adds a column bind to dataframe on left side

- binding rows --> use rbind instead of cbind

##Summarizing Data
- head(dataset, n=3)
- tail(dataset, n=1)
- summary(dataset)
- str(dataset) --> tells class of variable also 

- quantile(dataframe$columnname, no.rm=TRUE)
    - can pass through probability too --> quantile(xx$column, probs=c(0.5, 0.75, 0.9))
          - can look at different percentiles through this

```{r}
head(iris, n=3)

tail(iris, n=2)

summary(iris)

str(iris)

quantile(iris$Petal.Width, na.rm=TRUE)
quantile(iris$Petal.Width, probs=c(0.5, 0.75, 0.9))
quantile(iris$Petal.Width, probs=c(0.3, 0.75, 0.99))

```
#Summary Continues
- Can create a table --> table(df$column, useNA="ifany")
    - adds an additional column if any NAs
- default function does NOT tell us about NA

- 2D table --> pass two variables
    - table(df$col1, df$col2)
    
```{r}
table(iris$Sepal.Length, useNA = "ifany")
table(iris$Sepal.Length, iris$Sepal.Width) #sence of relationship between diff variables
```
#Check for missing values
- SUM is.na --> sum(is.na(df$column))
      - if 0 --> no NA
      
- ANY is.na --> any(is.na(df$column))
      - if any NA --> true
      
- ALL is.na --> all(is.na(df$column))
    - can also use for other conditions

```{r}
sum(is.na(iris$Sepal.Length))
any(is.na(iris$Sepal.Length))

all(iris$Sepal.Length > 1) # check to see if all have value of >1 
```
#Taking Sums
- Across columns or rows
    - colSums() --> can use with is.na to see check columns for missing values
    - rowSums() --> can use with all

```{r}
colSums(is.na(iris))
all(colSums(is.na(iris))==0) #all columns do not contain NA

```
#Finding specific details
- Finding everything equal to something:
    - table(df$column %in% c("xyz", "abc", "etc"))
        - similar to WHERE IN --> SQL
    - can also see the == command if don't want to check multiple items

```{r}
table(iris$Sepal.Length %in% c("1.2", "1.3", "5.1")) #how many fall in this list
```
#Subset data set using logic above
- use logic in subset
    - x[x$col %in% c("1", "2"), ] --> checking all rows 

```{r}
iris[iris$Sepal.Length %in% c("5.1", "5,2"),] #gets all rows with length of 5.1 or 5.2
```
#Summary Cross Tab
- summary(DF) 
- create DF using as.data.frame()

- Cross tab
    - use xtabs(...)
    - xt <- xtabs(Y ~ X1 + X2, data=DF)
        - Y is the value appearing in the tables; X1 and X2 are on the col/row labels

```{r}
data("UCBAdmissions")
DF <- as.data.frame(UCBAdmissions)
summary(DF)

xt <- xtabs(Freq ~ Gender + Admit, data=DF) #freq of each gender admitted
xt
```

#Flat Tables
- Cross tabs for larger variables
    - use xtabs --> can break down by . --> all
        - xtabs(Y ~., data=DF)

- If all lot of combos, there will be multipl tables. Can use ftable() to summarize easier

```{r}
warpbreaks$replicate <- rep(1:9, len = 54)
xt <- xtabs(breaks ~., data=warpbreaks)
#xt

ftable(xt)

```

#Data Set Siz
- object.size()
- print(object.size(data), units = "Mb) --> listing units

```{r}
object.size(iris)
print(object.size(iris), units ="Kb")
```