---
title: "Cleaning - Week 1"
author: "Syed Arif Hussain"
date: '2022-05-04'
output: html_document
---

```{r setup}
print("Let's do this!")
```

## Starting Week 1
Components of tidy data:
    - Raw data
    - tidy data set
    - Codebook describing each variable (metadata)
    - Explicit and exact recipe you used to go from 1 > 2 > 3 (steps)
    

## Downloading Files (R)
  - Need to know which directories to be working in: getwd() and setwd()
  - Relative vs. absolute paths:
      - Relative: setwd("./data") or setwd("../")
      - Absolute: setwd("/Users/Arif/one)
  - Windows need \\ not /

```{r cars}
setwd("C:/Users/Arif/Documents/GitHub/datasciencecoursera/New")
getwd()

file.exists("C:/Users/Arif/Documents/GitHub/datasciencecoursera") #check if directory exists
dir.create("C:/Users/Arif/Documents/GitHub/datasciencecoursera/New") #create a directory if none

#download.file() #for downloading file
```

#Download File Example
  - Important parameters: fileurl, destfile (directory), method
  - Download baltimore camera data
  - if URL starts with http or https (use "curl" method if on mac for this) should be ok


```{r data}
fileurl <- "https://www.arcgis.com/apps/dashboards/4b64b6e8c0014b6998d767fcf077bfaf"
download.file(fileurl, destfile = "./New/covid.csv")

list.files("./New")

```


##Reading Local Flat Files
  - header = true
  - sep = "," for CSV (read.table())
  - read.csv() puts header = TRUE automatically and sep = ","
  - quote="" means no quotes in data
  - na.strings: ser char that reps missing value
  - nrows
  - skip: number of lines skip before read

``` {r flat file}
covid_data <- read.csv("C:/Users/Arif/Documents/GitHub/datasciencecoursera/New/covid_data_cleaned.csv", header = TRUE)

head(covid_data)

summary(covid_data)

```

##Reading Excel
  - use library(xlsx) package as it's useful
     - can run: read.xlsx() and read.xlsx2()
          - sheetindex=1
  - reading specific rows or columns:
     - read.xlsx(..., colIndex = colIndex, rowIndex = rowIndex)
  - write.xlsx will write an excel file
      - library(XLConnect) and library(XLConnect vignette) both useful for Excel use


```{r Excel}
library(xlsx)

colIndex <- 1:3
rowIndex <- 1:4

```


##Readig XML
  - Extensible markup language
  - Components: 1) markup: label that gives structure to text, 2) actual text of document
  
  - Official tutorials: xml short and long
  
  
##Reading JSON
  - Java Script Object Notation
  - common format for APIs
  - Different syntax but similar to XML
  - Use libarary(jsonlite)
      - get file: fromJSON() function
      - save: toJSON()
  - json.org and wiki are good resources  
      
```{r JSON}
library(xml2)
#read_xml()

library(jsonlite)

iris2 <- toJSON(iris, pretty = TRUE) #taking dataset to JSON
iris3 <- fromJSON (iris2) #back to dataframe
head(iris2)
head(iris3)
```

##Data.Table Package
  - Inherits from data.frame
  - Faster than data.frame as written in C
  - Great for subsetting, grouping, updating
      - by=x (group by x variable)
  - New syntax
  - tables() #shows all tables in data 

  
  
```{r data.table}
library(data.table)

DF = data.frame(x=rnorm(9), y=rep(c("a", "b", "c"), each=3), z=rnorm(9))
head(DF)

DT = data.table(x=rnorm(9), y=rep(c("a", "b", "c"), each=3), z=rnorm(9))
head(DT)

DT[2,] #can subset same way
DT[DT$y=="a"]
DT[c(2,3)]
DF[c(2,3)]

DT[,list(mean(x), sum(z))]
DT[,w:=z^2]

DT[by=x]


```
